---
title: "Time Series Analysis of Stroke Incidence and Deaths"
author: "Adejumo Ridwan Suleiman"
date: "`r Sys.Date()`"
output: html_document
---

# Libraries
```{r}
library(tidyverse)
library(fpp2)
library(ggpubr)
```

# Data Cleaning and Manipulation
```{r}
#load stroke data
stroke <- read_csv("~/Time-Series-Analysis-of-Stroke/Stroke time.csv") |> 
  mutate(Date = lubridate::dmy(Date),
         Age_Categorized = 
           case_when(Age < 45 ~ "Less Than 45",
                     Age >= 45 & Age < 65 ~ "45-64",
                     Age >= 65 ~ "Greater or Equal To 65"),
         Occupation = case_when(Occupation == "Farmer" ~ "Skilled",
                                Occupation == "Dependent" ~ "Dependent",
                                Occupation == "Professional" ~ "Professional",
                                Occupation == "Skilled" ~ "Skilled"),
         MonthName = lubridate::month(Date,label = TRUE),
         Year = lubridate::year(Date)) |> 
  select(-`Month Name`)

stroke$Age_Categorized <- fct_relevel(stroke$Age_Categorized, c("Less Than 45","45-64","Greater or Equal To 65"))


# stroke incidence
incidence <- stroke |> 
  group_by(Date) |> 
  summarize(Incidence = n())

# stroke Deaths
deaths <- stroke |> 
  filter(Outcome == "Dead") |> 
  group_by(Date) |> 
  summarize(Deaths = n())

#Table of Incidence and Deaths
Inc_Deaths <- incidence |> 
  inner_join(deaths, by = "Date")

#Training Data 
Inc_Deaths_train <- Inc_Deaths |> 
  filter(Date <= "2019-12-01")

```




# Exploratory Data Analysis

## Incidence and Deaths of Stroke in Various Age groups from 2010-2021
```{r}
A <- stroke |> 
  group_by(Year, Age_Categorized) |> 
  summarize(Count = n()) |> 
  mutate_at(vars(Year),factor) |> 
  ggplot(aes(y = Year, x = Count,fill = Year)) +
  geom_col() +
  facet_wrap(Age_Categorized ~ .) +
  xlab("Number of Cases") +
  ggtitle("Incidence of Stroke in Various Age groups from 2010-2021") +
  theme(axis.title.y=element_blank(),
  axis.text.y=element_blank(),
  axis.ticks.y=element_blank(),
  legend.position = "none")

B <- stroke |> 
  filter(Outcome == "Dead") |> 
  group_by(Year, Age_Categorized) |> 
  summarize(Count = n()) |> 
  mutate_at(vars(Year),factor) |> 
  ggplot(aes(y = Year, x = Count,fill = Year)) +
  geom_col() +
  facet_wrap(Age_Categorized ~ .) +
  xlab("Number of Deaths") +
  ggtitle("Number of Deaths in Various Age groups from 2010-2021") +
  theme(axis.title.y=element_blank(),
    axis.text.y=element_blank(),
    axis.ticks.y=element_blank(),
    legend.key.size = unit(0.3, 'cm')) +
  guides(shape = guide_legend(override.aes = list(size = 0.5)))

ggarrange(A, B,nrow = 2)
```


## Incidence and Deaths of stroke in various Gender from 2010-2021
```{r}
C <- stroke |> 
  group_by(Year, Gender) |> 
  summarize(Count = n()) |> 
  mutate_at(vars(Year),factor) |> 
  ggplot(aes(y = Gender, x = Count,fill = Gender)) +
  geom_col() +
  facet_wrap(Year ~ .) +
  xlab("Number of Cases") +
  ggtitle("Incidence of Stroke in Various Gender from 2010-2021") +
  theme(axis.title.y=element_blank(),
  axis.text.y=element_blank(),
  axis.ticks.y=element_blank(),
  legend.position = "none")

D <- stroke |> 
  filter(Outcome == "Dead") |> 
  group_by(Year, Gender) |> 
  summarize(Count = n()) |> 
  mutate_at(vars(Year),factor) |> 
  ggplot(aes(y = Gender, x = Count,fill = Gender)) +
  geom_col() +
  facet_wrap(Year ~ .) +
  xlab("Number of Deaths") +
  ggtitle("Number of Deaths in Various Gender from 2010-2021") +
  theme(axis.title.y=element_blank(),
    axis.text.y=element_blank(),
    axis.ticks.y=element_blank(),
    legend.key.size = unit(0.3, 'cm')) +
  guides(shape = guide_legend(override.aes = list(size = 0.5)))

ggarrange(C, D,nrow = 2)
```



## Incidence and Deaths of stroke in various occupations from 2010-2021
```{r}
E <- stroke |> 
  group_by(Year, Occupation) |> 
  summarize(Count = n()) |> 
  mutate_at(vars(Year),factor) |> 
  ggplot(aes(y = Occupation, x = Count,fill = Occupation)) +
  geom_col() +
  facet_wrap(Year ~ .) +
  xlab("Number of Cases") +
  ggtitle("Incidence of Stroke in Various Occupations from 2010-2021") +
  theme(axis.title.y=element_blank(),
  axis.text.y=element_blank(),
  axis.ticks.y=element_blank(),
  legend.position = "none")

Fa <- stroke |> 
  filter(Outcome == "Dead") |> 
  group_by(Year, Occupation) |> 
  summarize(Count = n()) |> 
  mutate_at(vars(Year),factor) |> 
  ggplot(aes(y = Occupation, x = Count,fill = Occupation)) +
  geom_col() +
  facet_wrap(Year ~ .) +
  xlab("Number of Deaths") +
  ggtitle("Number of Deaths in Various Occupations from 2010-2021") +
  theme(axis.title.y=element_blank(),
    axis.text.y=element_blank(),
    axis.ticks.y=element_blank(),
    legend.key.size = unit(0.3, 'cm')) +
  guides(shape = guide_legend(override.aes = list(size = 0.5)))

ggarrange(E, Fa,nrow = 2)
```



## Table showing stroke Incidence
```{r}
stroke |> 
  group_by(Year, MonthName) |> 
  summarize(Count = n()) |> 
  mutate_at(vars(Year),factor) |> 
  pivot_wider(names_from = MonthName,
              values_from = Count,
              values_fill = 0) |> 
  write_csv(file = "incidence.csv")
```

## Table showing stroke deaths
```{r}
stroke |> 
  filter(Outcome == "Dead") |> 
  group_by(Year, MonthName) |> 
  summarize(Count = n()) |> 
  mutate_at(vars(Year),factor) |> 
  pivot_wider(names_from = MonthName,
              values_from = Count,
              values_fill = 0) |> 
  write_csv(file = "death.csv")
```


## Stroke Incidence by Months
```{r}
stroke |> 
  group_by(Year, MonthName) |> 
  summarize(Count = n()) |> 
  mutate_at(vars(Year), factor) |> 
  ggplot(aes(x = Year, y = Count,fill = Year)) +
  geom_col() +
  facet_wrap(MonthName ~ .) +
  theme(axis.title.x=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank()) +
  ylab("Number of Cases")+
  ggtitle("Monthly Stroke Incidence")
```

## Stroke Deaths by Months
```{r}
stroke |> 
  filter(Outcome == "Dead") |> 
  group_by(Year, MonthName) |> 
  summarize(Count = n()) |> 
  mutate_at(vars(Year), factor) |> 
  ggplot(aes(x = Year, y = Count,fill = Year)) +
  geom_col() +
  facet_wrap(MonthName ~ .) +
  ylab("Number of Deaths") +
  ggtitle("Monthly Stroke Deaths") +
  theme(axis.title.x=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank())
```



# Time Series Analysis of Incidence
```{r}
# Convert the Data to times series graphics
ts_train <- ts(Inc_Deaths_train[,c("Incidence","Deaths")],
                   frequency = 12,
                   start = c(2010, 01),
                   end = c(2019, 12))

```

```{r}
#stl decompostion
incidence_fit <- stl(ts_train[, "Incidence"], 
           s.window = "periodic",
           robust = TRUE)

# Decompositon plots
autoplot(incidence_fit) +
  ggtitle("STL decompostion of Stroke Incidence")
```

```{r}
#ARIMA modelling
set.seed(2022)
fit_arima_incidence <- auto.arima(ts_train[, "Incidence"],
                        stepwise = F,
                        approximation = F,
                        D = 1)
fit_arima_incidence
```

```{r}
#check residuals
checkresiduals(fit_arima_incidence)
```

```{r}
# Forecast in the next two years
fit_arima_incidence |> 
  forecast(23) |> 
  autoplot() +
  ylab("Number Of Cases") +
  xlab("Years")
```

```{r}
#forecast for the next 5 years
fit_arima_incidence |> 
  forecast(61) |> 
  autoplot() +
  ylab("Number Of Cases") +
  xlab("Years")
```


# Time Series Analysis of Deaths
```{r}
#stl decompostion
deaths_fit <- stl(ts_train[, "Deaths"], 
           s.window = "periodic",
           robust = TRUE)

# Decompositon plots
autoplot(deaths_fit) +
  ggtitle("STL decompostion of Deaths")
```
```{r}
#ARIMA modelling
deaths_fit_arima <- auto.arima(ts_train[, "Deaths"],
                        stepwise = F,
                        approximation = F,
                        D = 1)
deaths_fit_arima
```

```{r}
checkresiduals(deaths_fit_arima)
```

```{r}
#forecast for 2 years
deaths_fit_arima |> 
  forecast(23) |> 
  autoplot() +
  ylab("Number of Deaths") +
  xlab("Year")
```
```{r}
#forecast for the next 5 years
deaths_fit_arima |> 
  forecast(61) |> 
  autoplot() +
  ylab("Number of Deaths") +
  xlab("Year")
```
